---
title: 'Assignment 2: Data visualization'
author: "Tamas Nagy"
output: html_document
editor_options: 
  chunk_output_type: console
---

You will have to create 3 plots based on the datasets and instructions detailed below. You will find the plots themeselves in the `assignments/assignment_2_plots`. Your task is to write the code that will reproduce the plots as closely as possible.

# Skills needed to solve this assignment

-   Using R and RStudio, reading data
-   Reporting using RMarkdown
-   Using Git and Github (for submitting the task)
-   Data manipulation (e.g. dplyr, tidyr), and working with factors (forcats)
-   Data visuzlization (ggplot2)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(tidytuesdayR)
library(viridis)
library(forcats)
```

## Task 1: Climbing expeditions

The 2020-09-22 TidyTueday datasets are about climbing expeditions. From the three datasets, use the "expeditions". Reproduce the plot below! Notice a few things:

-   Use `forcats::fct_lump()` to get the 15 most frequent peaks, and drop the "Other" category.
-   The bars are ordered by the sum of all expeditions (use `fct_reorder()`).
-   The bar colors use the viridis palette and light theme.

```{r}
expeditions <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2020/2020-09-22/expeditions.csv')

# Variables used: count/summarize the number of expeditions by the peak name and categorize by season during the plotting

# creating a data frame for the first plot
# Grouping the first 15 peak, dropping the others by filtering the data frame we use for the plot

first_plot <- expeditions %>%
  mutate(
    peak_name = fct_lump(peak_name, n = 15, other_level = "Other")
  )

first_plot <- first_plot %>%
  filter(peak_name != "Other")
# Counting the number of 
first_plot <- first_plot %>%
  count(peak_name, season, name = "n") %>%
  group_by(peak_name) %>%
  mutate(total = sum(n)) %>%
  ungroup() %>%
  mutate(peak_name = fct_reorder(peak_name, total))

```

## Plotting the 1st graph
```{r}
firstplot <- ggplot(first_plot, aes(x = peak_name, y = n, fill = season)) +
  geom_col(width = 0.8) +
  coord_flip() +
  scale_fill_viridis_d() +
  labs(
    x = NULL,
    y = "Number of Expeditions",
    title = "The 15 most popular peaks stacked by season of expedition",
    fill = "season"
  ) +
  theme_light() + 
  theme(legend.position = "bottom",
        text = element_text(size = 16),
        plot.title = element_text(size = 17))
```

## Saving the plot to check for differences
```{r}
# ggsave("first_replicated_plot.png", firstplot, width = 16, height = 9, dpi = 400)
# getwd()
# Plot checked, looks similar
```

## Task 2: PhDs awarded

The 2019-02-19 TidyTueday dataset is about phd-s awarded by year and field. There is only one dataset, it is called `phd_by_field`. Reproduce the plot below!

Notes:

-   First you have to aggregate the data to count all phd-s by broad fields.
-   To make the x axis breaks pretty, use `scales::pretty_breaks()`, to make the y axis labels comma formatted, use `scales::comma_format()`.
-   The line size is 1.2, the colors are from the brewer "Dark2" palette. The theme is set to minimal.

```{r}
phd_field <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2019/2019-02-19/phd_by_field.csv")

# aggregate the data to count all board field
# x axis: years
# using line plot

phd_plot_data <- phd_field %>%
  group_by(year, broad_field) %>%
  summarise(total_phds = sum(n_phds, na.rm = TRUE)) %>%
  ungroup()

phd_plot <- ggplot(phd_plot_data, aes(x = year, y = total_phds, color = broad_field)) +
  geom_line(size = 1.2) +
  scale_color_brewer(palette = "Dark2") +
  scale_x_continuous(breaks = scales::pretty_breaks()) +
  scale_y_continuous(labels = scales::comma_format()) +
  labs(
    x = NULL,
    y = NULL,
    color = "Broad field",
    title = "Number of awarded Ph.D.-s in the US by year"
  ) +
  theme_minimal()
```

## Plot checking via printing
```{r}
# ggsave("second_replicated_plot.png", phd_plot)
```

## Task 3: Commute in the US

The 2019-11-05 TidyTueday dataset is about commuting to work in each city in the US by bike or on foot. There is only one dataset, it is called `commute`. Reproduce the plot below!

Notes:

-   First you have to aggregate the data to count all commutes by state.
-   Both axis scales are log transformed and the labels comma formatted, using `scales::comma_format()`
-   The point size is 2, . The theme is set to light.

```{r}
commute_mode <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2019/2019-11-05/commute.csv")

# grouping the data and counting the state and mode combinations
commute_plot_data <- commute_mode %>%
  group_by(state, mode, state_region, state_abb) %>%
  summarise(total = sum(n, na.rm = TRUE)) %>%
  ungroup()

```
## Change from long to wide format

```{r}
commute_plot_data <- commute_plot_data %>%
  pivot_wider(names_from = mode, values_from = total) %>%
  filter(!is.na(Walk), !is.na(Bike), !is.na(state_abb))

# The original plot contained NA-s as well, but they were hard to notice and i decided not to include them
```

## Plotting
```{r}
walkbike_plot <- ggplot(commute_plot_data, aes(x = Walk, y = Bike, color = state_region)) +
  geom_point(size = 2) +
  geom_text(aes(label = state_abb), color = "black", size = 3.5, fontface = "bold") +
  scale_x_log10(labels = scales::comma_format()) +
  scale_y_log10(labels = scales::comma_format()) +
  labs(
    x = "Number of people walking to work (log N)",
    y = "Number of people biking to work (log N)",
    color = "State region",
    title = "Number of people walking vs. biking to work in each USA state"
  ) +
  theme_light() +
    theme(text = element_text(size = 14)
  )
```
## Saving to check

```{r}
# ggsave("third_plot_replicate.png", walkbike_plot)
# checked, it looks similar, just the starting point of the y scale is higher now, as we eliminated the NA-s
```

